A thermodynamic view of dusty protoplanetary disks

### 2017-05-08 model.tex
extra intro section

The total gravitational potential is $\Phi_\mathrm{tot}=\Phi +
\psi$. Here $\Phi$ originates from a star of mass $M_*$, 
\begin{align}\label{thin_disk_potential}
  \Phi(r,z) =-\frac{GM_*}{\sqrt{r^2 + z^2}}\simeq
  -\frac{GM_*}{r}\left(1 - \frac{z^2}{2r^2}\right),  
\end{align}
where $(r,\phi, z)$ are cylindrical co-ordinates centered on the star.   
The second equality in Eq. \ref{thin_disk_potential} is the 
thin-disk approximation for the stellar potential, appropriate for
$|z|\ll r$. We shall adopt this approximate potential in order to
obtain explicit expressions for disk equilibria (\S\ref{eqm}).

The disk's own gravitational potential $\psi$ is included for
completeness, but we will mostly neglect it unless stated
otherwise. In such cases $\psi$   
satisfies  
\begin{align}
  \nabla^2\psi = 4 \pi G \rho,
\end{align}
where $G$ is the gravitational constant.

>>>>

The total gravitational potential is $\Phi_\mathrm{tot}=\Phi +
\psi$. Here $\Phi$ originates from a star of mass $M_*$, 
\begin{align}\label{thin_disk_potential}
  \Phi(r,z) =-\frac{GM_*}{\sqrt{r^2 + z^2}}\simeq
  -\frac{GM_*}{r}\left(1 - \frac{z^2}{2r^2}\right),  
\end{align}
where $(r,\phi, z)$ are cylindrical co-ordinates centered on the star, and 
$G$ is the gravitational constant. The second equality is the  
thin-disk approximation for the stellar potential, appropriate for
$|z|\ll r$. We shall adopt this approximate potential in order to
obtain explicit expressions for disk equilibria (\S\ref{eqm}).

The disk's own gravitational potential $\psi$ is included for
completeness, but we will mostly neglect it unless stated
otherwise. In such cases $\psi$ satisfies the Poisson equation 
\begin{align}
  \nabla^2\psi = 4 \pi G \rho.
\end{align}

### 2017-05-08 paper.tex
extra intro section

This paper is organized as follows. We first
explain in \S\ref{grow_osc} a general result of fluid dynamics that
underpins the rest of our study. We describe how 
a pressure-density phase lag in fluids can lead to 
instability. Consequently, an analogy between dusty gas and  
pure hydrodynamics allow us to interpret dusty-gas instabilities 
within a thermodynamics framework.

@@

>>>>

This paper is organized as follows. In \S\ref{key_idea} we first
describe, without derivation, the analogy between dusty gas and    
ideal hydrodynamics. We then discuss in 
\S\ref{grow_osc} a general result of fluid dynamics that
underpins the bulk of our study: intabilities due to 
pressure-density phase lags. This means that dusty-gas phenomena, such
as SI, have thermodynamical interpretations.

@@

\section{Dust-free modeling of dusty protoplanetary disks}\label{key_idea}
{\bf is this section needed?} 
The key concept we introduce in this work is the description
of dusty protoplanetary disks \emph{without} having dust as an
explicit variable. This is best illustrated by writing down our model  
equations that describe a mixture of isothermal gas and small
dust particles: 
\begin{align}  
  &\frac{\p \rho}{\p t} + \nabla\cdot\left(\rho\bm{v}\right)= 0, \label{mass0}\\ 
  &\frac{\p\bm{v}}{\p t} + \bm{v}\cdot\nabla\bm{v} = -
  \frac{1}{\rho}\nabla  P - \nabla \Phi_*,\label{mom0}\\ 
 & \frac{\p P}{\p t} + \bm{v}\cdot\nabla P 
  = - P \nabla\cdot\bm{v}
                + \mathcal{C}(\rho, P; \tstop)\label{en0}, 
\end{align}
where $\rho=\rhod + \rhog$ is the total density, $P$ is the pressure,
$\Phi_*$ is the stellar gravitational potential,   
and $\mathcal{C}$ represent effects of dust-gas drag, which
is paremterized by the stopping time $\tstop$.

Eq. \ref{mass0}---\ref{en0} appear identical to adiabatic gas dynamics, even  
though they physically represents a two-phase, dusty-gas. The main
difference is the term $\mathcal{C}$. Here it does not represent
physical cooling (as the system is isothermal). Instead, dust-gas drag
causes particles to drift, hence by momentum conservation a gas drift
in the opposite direction, which changes the pressure distribution.

Since the above equations are similar to ideal gas dynamics, it 
means that many of the results derived for ideal hydrodynamics can be  
generalized to dusty disks without additional effort. This includes 
the important effect of phase lags between density and pressure in
fluids, discussed next.

### 2017-05-08 summary.tex
extra intro section

\subsubsection{Convective overstability}
The `convective overstability' (ConO) operates in pure gas, non-adiabatic disks 
when there is a radial buoyancy frequency such that $N_r^2\equiv
F_r\p_rS/C_P  < 0 $ and cooling times $t_\mathrm{cool}\sim \OmK^{-1}$. This
combination leads to growing epicycles
\citep{klahr14,lyra14,latter16}.     
%with 
%growth rate $\propto  -N_r^2$ 
% Indeed, the effective radial buoyancy frequency in our dusty disk 
% with negligble background density gradient is unstable, 
% \begin{align}
%   N_r^2 \equiv -\frac{1}{\rho}\frac{\p P}{\p r}\frac{\p S_\mathrm{eff}}{\p r} =
%   -\frac{\rho}{P}F_r^2<0.  
% \end{align}

@@

Now consider a strictly isothermal dusty disk where the relevant 
entropy is dust-induced, $\seff =-\ln{(1+\rhod/\rhog)}$. 
%The dusty/adiabatic gas
%analogy implies the replacement $S\to\seff = -\ln{(1+\rhod/\rhog)}$ in
%$N_r$. 
Since $F_r\propto -\p_rP > 0$ in typical disk models, ConO would
require $\p_r\seff<0$, implying the dust-to-gas ratio increases
\emph{outwards}. This might be realized at special locations in
protoplanetary disks.

@@

Isothermal disks have $\tcool=0$, so the pure gas ConO cannot
exist. However, we have shown that dust-gas drag provides an effective
energy source/sink to the mixture (\S\ref{energy_analogy}). It would be
interesting to explore whether or not dust-gas drag can provide the 
required `cooling' to enable a dusty version of the convective 
overstability.

@@

On the other hand, we have show that dust-loading an isothermal disk
also induces an effective buoyancy. % This effect is maximized when dust is
% perfectly coupled to the gas
. It is thus natural to ask whether or not
the required adiabatic conditions for ZVI can be realized through
dust-loading, and thus produce a `dusty ZVI'. This may allow the ZVI
to operate at $\gtrsim 1$AU in PPDs.

>>>>

\subsubsection{Convective overstability}
The `convective overstability' (ConO) was discovered in pure gas, non-adiabatic, unstratified disk models    
where the radial buoyancy frequency is such that $N_r^2\equiv
F_r\p_rS/C_P  < 0 $ and cooling times $t_\mathrm{cool}\sim \OmK^{-1}$. This
combination leads to growing epicycles
\citep{klahr14,lyra14,latter16}.     
%with 
%growth rate $\propto  -N_r^2$ 
% Indeed, the effective radial buoyancy frequency in our dusty disk 
% with negligble background density gradient is unstable, 
% \begin{align}
%   N_r^2 \equiv -\frac{1}{\rho}\frac{\p P}{\p r}\frac{\p S_\mathrm{eff}}{\p r} =
%   -\frac{\rho}{P}F_r^2<0.  
% \end{align}

@@

Now consider a strictly isothermal dusty disk where the relevant 
entropy is dust-induced, $\seff =-\ln{(1+\rhod/\rhog)}$. 
%The dusty/adiabatic gas
%analogy implies the replacement $S\to\seff = -\ln{(1+\rhod/\rhog)}$ in
%$N_r$. 
Since $F_r\propto -\p_rP > 0$ in typical disk models, ConO would
require $\p_r\seff<0$, implying the dust-to-gas ratio increases
\emph{outwards}. This might be realized at special radial locations in 
protoplanetary disks, such as planet gaps. Dust-settling itself may
also lead to the mid-plane dust-to-gas ratio to increase outwards
\citep{takeuchi02}.

@@

Isothermal disks have $\tcool=0$, so the pure gas ConO cannot
exist. However, we have shown that dust-gas drag provides an effective
energy source/sink to the mixture (\S\ref{energy_analogy}). It would be
interesting to explore whether or not dust-gas drag can provide the 
required `cooling' to enable a `dusty convective 
overstability'.

@@

On the other hand, we have show that dust-loading an isothermal disk
also induces an effective buoyancy. % This effect is maximized when dust is
% perfectly coupled to the gas
. It is thus natural to ask whether or not
the required adiabatic conditions for ZVI can be realized through
dust-loading, and thus produce `dusty ZVI'. This may allow the zombie
vortices to exist at $\gtrsim 1$AU in PPDs.
